Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2018-12-31 astronum_2018/astronum-xrb.tex
add e-mail
some figs
updated version
add note about GPUs

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{xrb}
\caption{\label{fig:xrb} Helium flame spreading across the surface of
  a neutron star.  Shown are temperature (top), the mean molecular
  weight of the ash (second), the nuclear energy generation rate
  (third), and the velocity out of the simulation plane (bottom).  The
  out-of-plane velocity contours indicate where the Coriolis force has
  created a hurricane-like wind structure which confines the flame,
  leading to an inclined front. It is at such an interface that the
  temperature and burning rate are highest.  }
\end{figure}

@@

Machine architectures are evolving, with heterogeneous architectures
becoming more common.  Our recent development in \castro\ has focused
on moving the hydrodynamics to GPUs~\cite{astronum:2017}.  The XRB
problem is an ideal case for this, since hydrodynamics with constant
gravity (for plane-parallel domains) ports in a straightforward manner
to GPUs.  The current GPU version of the \castro\ hydrodynamics solver
is about $10\times$ faster using GPUs on a node than the entire node
of CPU cores.  We see similar speed ups with reaction networks, with
these GPU ports under active development.  This means a GPU
calculation could lower the above costs by a factor of 10.  GPU
offloading of the reaction networks will also allow us to explore
larger networks and more detailed nuclear physics.

@@

As an example of the improvements in coupling between hydro
and reactions, Figure~\ref{fig:sdc} shows the mass fraction of helium
behind a detonation over the course of 2 timesteps.  For this test,
a 19 isotope reaction network was used.  The points
represent individual calls to the righthand side function of our
reaction network, and the density of
points indicates how hard the integrator is working.  We use the same
tolerances for both the Strang-split method and the SDC method.  For
the SDC method, we are using a second-order accurate method based on
\cite{SDC-old}.  We predict a time-centered advection term,
$\Adv{\Uc}^{n+1/2}_i$ using standard unsplit Godunov methods, but
explicitly include a reactive source term in the tracing of the
interface states, ${\bf R}(\Uc_i)$.  We can then solve the reactive
system, using this advective term as a piecewise-constant-in-time
source
\begin{equation}
\label{eq:syst}
\frac{d{\Uc_i}}{dt} = -\Adv{\Uc}^{n+1/2}_i + {\bf R}(\Uc_i)
\end{equation}
This system can be integrated with standard ODE methods.  To achieve
second-order accuracy, we need to iterate, using the updated $\Uc$ to
create a reactive source included in the predictor for the advective
update, and then reintegrate the system.  The result of this iteration
is that the advection sees the effects of reactions over the timestep
and the reactions see the effects of advection.  For the zone tracked
in Figure~\ref{fig:sdc}, each SDC iteration called the righthand side
function of the network 2 times less than the Strang case.  This
savings is dependent on the network and thermodynamic state, but our
tests thus far show that the SDC methods need fewer network calls to
integrate a timestep.  We also clearly see how the Strang system
evolves in a discontinuous fashion.  We are continuing to develop this
new integration strategy, with full fourth-order in space and time
reacting hydrodynamics methods to follow shortly.  The XRB simulations
are our target application.

>>>>

\ead{michael.zingale@stonybrook.edu}

@@

\begin{figure}[t]
\centering
\includegraphics[width=\linewidth]{xrb}
\caption{\label{fig:xrb} Helium flame spreading across the surface of
  a neutron star.  Shown are temperature (top), the mean molecular
  weight of the ash (second), the nuclear energy generation rate
  (third), and the velocity out of the simulation plane (bottom).  The
  out-of-plane velocity contours indicate where the Coriolis force has
  created a hurricane-like wind structure which confines the flame,
  leading to an inclined front. It is at such an interface that the
  temperature and burning rate are highest.  }
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.9\linewidth]{xrb_profiles}
\caption{\label{fig:xrb_profiles} Vertical profiles at several laterial distances
  from the vertical axis through the plots shown in Figure~\ref{fig:xrb}.}
\end{figure}

@@

Machine architectures are evolving, with heterogeneous architectures
becoming more common.  Our recent development in \castro\ has focused
on moving the hydrodynamics to GPUs~\cite{astronum:2017}.  The XRB
problem is an ideal case for this, since hydrodynamics with constant
gravity (for plane-parallel domains) ports in a straightforward manner
to GPUs.  The current GPU version of the \castro\ hydrodynamics solver
is about $10\times$ faster using GPUs on a node than the entire node
of CPU cores (comparing 4 NVIDIA Volta GPUs to 20 IBM Power9 cores).
We see similar speed ups with reaction networks, with these GPU ports
under active development.  This means a GPU calculation could lower
the above costs by a factor of 10.  GPU offloading of the reaction
networks will also allow us to explore larger networks and more
detailed nuclear physics.

@@

As an example of the improvements in coupling between hydro
and reactions, Figure~\ref{fig:sdc} shows the mass fraction of helium
behind a detonation over the course of 2 timesteps.  For this test,
a 19 isotope reaction network was used.  The points
represent individual calls to the righthand side function of our
reaction network, and the density of
points indicates how hard the integrator is working.  We use the same
tolerances for both the Strang-split method and the SDC method.  For
the SDC method, we are using a second-order accurate method based on
\cite{SDC-old}.  We predict a time-centered advection term,
$\Adv{\Uc}^{n+1/2}_i$ using standard unsplit Godunov methods, but
explicitly include a reactive source term in the tracing of the
interface states, ${\bf R}(\Uc_i)$.  We can then solve the reactive
system, using this advective term as a piecewise-constant-in-time
source
\begin{equation}
\label{eq:syst}
\frac{d{\Uc_i}}{dt} = -\Adv{\Uc}^{n+1/2}_i + {\bf R}(\Uc_i)
\end{equation}
This system can be integrated with standard ODE methods.  To achieve
second-order accuracy, we need to iterate, using the updated $\Uc$ to
create a reactive source included in the predictor for the advective
update, and then reintegrate the system.  The result of this iteration
is that the advection sees the effects of reactions over the timestep
and the reactions see the effects of advection.  For the zone tracked
in Figure~\ref{fig:sdc}, each SDC iteration called the righthand side
function of the network 2 times less than the Strang case.  This
savings is dependent on the network and thermodynamic state, but our
tests thus far show that the SDC methods need fewer network calls to
integrate a timestep.  We also clearly see how the Strang system
evolves in a discontinuous fashion.  We are continuing to develop this
new integration strategy, with full fourth-order in space and time
reacting hydrodynamics methods to follow shortly.  The XRB simulations
are our target application.  We note that there has been some
development in the community on implicit time-integration methods for
compressible flow, but for the XRB problem, the wide range of Mach
number seen in the domain (exceeding unity near the top of the
atmosphere, ahead of the flame) removes much of the advantages of
implicit timestepping, so we focus on explicit methods.
